# Inputs {#inputs}

## Input functions

Inputs are ways that users can communicate information to the Shiny app. Explore some different input types in the embedded app below before you read about how to set up each type.

```{r input-demo-app, eval = TRUE, echo=FALSE, fig.cap="Input Demo App. You can also access this app with `shinyintro::app(\"input_demo\")` or view it in a separate tab with the [showcase interface](https://shiny.psy.gla.ac.uk/debruine/input_demo/){target=\"_blank\"}."}
knitr::include_app("https://shiny.psy.gla.ac.uk/debruine/input_demo/",
  height = "800px")
```

### textInput

`r func("textInput")` creates a one-line box for short text input. The first `r glossary("argument")`, `r arg("inputId")` (the argument name is usually omitted), needs to be a unique `r glossary("string")` that you cannot use for another input or output in this app.

```{r}
demo_text <- 
  textInput("demo_text", 
            label = "Name", 
            value = "", 
            width = "100%",
            placeholder = "Your Name")
```

::: {.try}
Experiment with the values of `r arg("label")`, `r arg("value")`, `r arg("width")`, and `r arg("placeholder")` to see what they do.
:::

### textAreaInput

`r func("textAreaInput")` creates a multi-line box for longer text input.

```{r}
demo_textarea <- 
  textAreaInput("demo_textarea", 
                label = "Biography", 
                value = "",
                width = "100%",
                rows = 5, 
                placeholder = "Tell us something interesting about you.")
```

What is the `r arg("inputId")` of the widget above?
`r mcq(c("textAreaInput", answer="demo_textarea", "Biography"))`

### selectInput

`r func("selectInput")` creates a drop-down menu. Set the first `r arg("choice")` to `""` to default to `NA`. If your choices are a named `r glossary("list")` or `r glossary("vector")`, the names are what is shown and the values are what is recorded. If the choices aren't named, the displayed and recorded values are the same.

```{r}
demo_select <- 
  selectInput("demo_select", 
              label = "Do you like Shiny?", 
              choices = list("", 
                             "Yes, I do" = "y", 
                             "No, I don't" = "n"),
              selected = NULL,
              width = "100%")
```

If you set `r arg("multiple")` to `TRUE`, you can also make a select where users can choose multiple options.

```{r}
genders <- list( # no blank needed
  "Non-binary" = "nb",
  "Male" = "m",
  "Female" = "f",
  "Agender" = "a",
  "Gender Fluid" = "gf"
)

demo_select_multi <- 
  selectInput("demo_select2", 
              label = "Gender (select all that apply)", 
              choices = genders,
              selected = NULL,
              multiple = TRUE, 
              selectize = FALSE,
              size = 5)
```

### checkboxGroupInput

However, this interface almost always looks better with `r func("checkboxGroupInput")`. 

```{r}
demo_cbgi <-
  checkboxGroupInput("demo_cbgi",
                     label = "Gender (select all that apply)",
                     choices = genders)
```

How can you get the checkboxes to display horizontally instead of vertically?  
`r mcq(c("display = 'horizontal'", "class = 'horiz'", answer = "inline = TRUE", "class = 'shiny-input-container-inline'"))`


### checkboxInput

You can also make a single checkbox with `r func("checkboxInput")`. The `r arg("value")` is `TRUE` when checked and `FALSE` when not.

```{r}
demo_cb <- checkboxInput("demo_cb",
                         label = "I love R",
                         value = TRUE)
```

`r func("sliderInput")` allows you to choose numbers between a `r arg("min")` and `r arg("max")` value.

```{r}
demo_slider <- sliderInput("demo_slider",
                           label = "Age",
                           min = 0,
                           max = 100,
                           value = 0,
                           step = 1,
                           width = "100%")
```

::: {.try}
What happens if you change `r arg("value")` or `r arg("step")`? Try changing `r arg("value")` to `c(10, 20)`.
:::

### radioButtons

If you want users to only be able to choose one options and there are a small number of short options, `r func("radioButton")` is a good interface. 

```{r}
demo_radio <- radioButtons("demo_radio",
                           label = "Choose one",
                           choices = c("Cats", "Dogs"),
                           selected = character(0),
                           inline = TRUE)
```

::: {.warning}
Radio buttons default to selecting the first item unless you set `r arg("selected")` to a choice value or `character(0)` to start with no selection.
:::

### dateInput

I find the date interface a little clunky, but that might just be because I have to click the back button on the year interface 44 time to find my birthdate. However, it also allows you to type in a date following the format that you can set.

```{r}
demo_date <- dateInput("demo_date",
                       label = "What is your birth date?",
                       min = "1900-01-01",
                       max = Sys.Date(),
                       format = "yyyy-mm-dd",
                       startview = "year")
```

::: {.info}
IMHO, the default of `"yyyy-mm-dd"` is the best because it sorts into chronological order. Don't let me catch you storing dates like `"m/d/yyyy"`. 
:::

What would you set `r arg("format")` to in order to display dates like "Sunday July 4, 2021"?  
`r mcq(c("D M d, Y", answer = "DD MM d, yyyy", "DAY MONTH day, YEAR", "D MM dd, yyyy"))`

(don't actually do this)


### fileInput

Users can upload one or more files with `r func("fileInput")`. The argument `r arg("acccept")` lets you limit this to certain file types, but some browsers can bypass this requirement, so it's not fool-proof.

```{r}
demo_file <- fileInput("demo_file",
                       label = "Upload a data table",
                       multiple = FALSE,
                       accept = c(".csv", ".tsv"),
                       buttonLabel = "Upload")
```

What would you set `r arg("accept")` to to accept any image file?  
`r mcq(c(answer = "image/*",".jpg", "jpg", "images", ".img"))`


## Setting inputs programatically

Sometimes you need to change the value of an input with code, such as when resetting a questionnaire or in response to an answer on another item. The following code resets all of the inputs above.

```{r}
updateTextInput(session, "demo_text", value = "")
updateTextAreaInput(session, "demo_textarea", value = "")
updateSelectInput(session, "demo_select", selected = "")
updateCheckboxGroupInput(session, "demo_cbgi", selected = character(0))
updateCheckboxInput(session, "demo_cb", value = TRUE)
updateRadioButtons(session, "demo_radio", selected = character(0))
updateSliderInput(session, "demo_slider", value = 0)
updateDateInput(session, "demo_date", value = NULL)
```

::: {.warning}
Note that select inputs, checkbox groups, and radio buttons use the argument `selected` and not `value`. If you want to set all the values in a checkbox group or radio button group to unchecked, set `selected = character(0)`.
:::

## Exercises {#exercises-inputs}

### Pets

Create an interface that gets people to rate the following pets on a 9-point scale. You can use any option labels or input type you like.

* Dogs ðŸ•  
* Cats ðŸˆ  
* Birds ðŸ¦œ  
* Fish ðŸ   
* Mice ðŸ  
* Hedgehogs ðŸ¦”  
* Snakes ðŸ

### Questionnaire

In the app you're developing, add a tab for a questionnaire that you're interested in and set up the appropriate inputs. 

### Updating

Add a "reset" button to your questionnaire tab and write the `r func("server")` code to reset all of its inputs.
