# Outputs {#outputs}

Output are ways that the Shiny app can dynamically display information to the user. In the user interface (`r glossary("UI")`), you create outputs with IDs that you reference in an associated rendering function inside the `r glossary("server")` function.

Explore some different output types in the embedded app below before you read about how to set up each type.

```{r output-demo-app, eval=TRUE, echo=FALSE, fig.cap="Output Demo App. You can also access this app with `shinyintro::app(\"output_demo\")`or view it in a separate tab with the [showcase interface](https://shiny.psy.gla.ac.uk/debruine/output_demo/){target=\"_blank\"}."}
if (is_html)
  knitr::include_app("https://shiny.psy.gla.ac.uk/debruine/output_demo/", height = "670px")
```


## Text

`r func("textOutput")` defaults to text inside a generic `<span>` or `<div>`, but you can use a different `r glossary("element-html", "element")` with the `r arg("container")` argument.

```{r}
# in the UI function
textOutput("demo_text", container = tags$h3)
```

`r func("renderText")` replaces the text of the linked element with its returned string.

```{r}
# in the server function
output$demo_text <- renderText({
    sprintf("Plot of %s", input$y)
})
```

If you use `r func("verbatimTextOutput")` in the UI (no change to the render function), it will show the output in a fixed-width font. This can be good for code or text you want the user to copy.

```{r}
# in the UI function
verbatimTextOutput("demo_verbatim")

# in the server function
output$demo_verbatim <- renderText({
  code <-
    "ggplot(iris, aes(x = Species, y = %s, color = Species)) +
    geom_violin(show.legend = FALSE) +
    stat_summary(show.legend = FALSE)"
  
  sprintf(code, input$y)
})
```


## Plots

`r func("plotOutput")` displays plots made with the `r glossary("base R")` plotting functions (e.g., `r func("plot")`, `r func("hist")`) or `r pkg("ggplot2")` functions.

```{r}
# in the UI function
plotOutput("demo_plot", width = "500px", height="300px")
```

What is the default value for `r arg("width")`?  
`r mcq(c(answer = "100%", "400px", "400", "5in", "7in"))`  
What is the default value for `r arg("height")`?  
`r mcq(c("100%", answer = "400px", "400", "5in", "7in"))`


```{r}
# in the server function
output$demo_plot <- renderPlot({
  ggplot(iris, aes(x = Species, y = .data[[input$y]], color = Species)) +
    geom_violin(show.legend = FALSE) +
    stat_summary(show.legend = FALSE) +
    ylab(input$y)
})
```

::: {.warning data-latex=""}
If you want to create dynamic plots that change with input, note how you need to use `y = .data[[input$y]]` inside `r func("aes")`, instead of just `y = input$y`.
:::

## Images

`r func("imageOutput")` takes the same arguments as `r func("plotOutput")`. You can leave `r arg("width")` and `r arg("height")` as their defaults if you are going to set those values in the render function.

```{r}
# in the UI function
imageOutput("demo_image")
```

`r func("renderImage")` needs to return a named list with at least an `r arg("src")` with the image path. You can also set the `r arg("width")` and `r arg("height")` (numeric values are in pixels), `r arg("class")` and `r arg("alt")` (the alt-text for screen readers).

```{r}
# in the server function
output$demo_image <- renderImage({
    list(src = "images/flower.jpg",
         width = 100,
         height = 100,
         alt = "A flower")
}, deleteFile = FALSE)
```

::: {.warning data-latex=""}
The `r arg("deleteFile")` argument is currently optional, but triggers periodic warnings that it won't be optional in the future. You should set it to `r dt(TRUE)` if you're making a temporary file (this stops unneeded plots using memory) and `r dt(FALSE)` if you're referencing a file you previously saved.
:::

## Tables

Display a table using `r func("tableOutput")`.

```{r}
# in the UI function
tableOutput("demo_table")
```

This is paired with `r func("renderTable")`, which makes a table out of any data frame it returns.

```{r}
# in the server function
output$demo_table <- renderTable({
  iris %>%
    group_by(Species) %>%
    summarise(mean = mean(.data[[input$y]]),
              sd = sd(.data[[input$y]]))
})
```

::: {.warning data-latex=""}
Note how you need to use `.data[[input$y]]` inside `r func("dplyr::summarise")`, instead of just `input$y` to dynamically choose which variable to summarise.
:::

### Data Tables

If you have a long table to show, or one that you want users to be able to sort or search, use `r func("DT::dataTableOutput")` (or its synonym `r func("DTOutput")`). 

The basic `r pkg("shiny")` package has `r func("dataTableOutput")` and `r func("renderDataTable")` functions, but they can be buggy. The versions in the `r pkg("DT")` package are better and have [many additional functions](https://rstudio.github.io/DT/){target="_blank"}, so I use those. 

```{r}
# in the UI function
DT::dataTableOutput("demo_datatable",
                    width = "50%",
                    height = "auto")
```

The paired render function is `r func("renderDataTable")` (or its synonym `r func("renderDT")`). You can customise data tables in many ways, but we'll stick with a basic example here that limits the number of rows shown at once to `r dt(10)`.

```{r}
# in the server function
output$demo_datatable <- DT::renderDataTable({
    iris
}, options = list(pageLength = 10))
```

::: {.info data-latex=""}
You can use the `r pkg("DT")` synonyms to make sure you're not accidentally using the `r pkg("shiny")` versions, which don't have the same options.
:::

## Dynamic HTML

If you want to dynamically create parts of the UI, you can use `r func("uiOutput")`. 

```{r}
# in the UI function
uiOutput("demo_ui")
```

You can create the UI using `r func("renderUI")` to return HTML created using the input functions we learned about in Section\ \@ref(inputs), the [tag functions](#tags), or HTML that you write yourself (as an argument to `r func("HTML")`).

```{r}
# in the server function
output$demo_ui <- renderUI({
  cols <- names(iris)[1:4]
  selectInput("y", "Column to plot", cols, "Sepal.Length")
})
```

::: {.info data-latex=""}
The function `r func("htmlOutput")` is a synonym for `r func("uiOutput")`, so you might see that in some code examples, but I use `r func("uiOutput")` to make the connection with `r func("renderUI")` clearer, since there is no `r func("renderHTML")`.
:::

## Further Resources {#reosurces-outputs}

* [Mastering Shiny Section 2.3](https://mastering-shiny.org/basic-ui.html#outputs){target="_blank"}
* [RStudio Shiny Tutorials](https://vimeo.com/rstudioinc/review/131218530/212d8a5a7a/#t=10m41s){target="_blank"}
* [DT (data tables)](https://rstudio.github.io/DT/){target="_blank"}

## Exercises {#exercises-outputs}

### Modify the demo {-}

Clone the "output_demo" app and modify it to use a different dataset.

### Pets {-}

Add outputs and appropriate render functions to show a plot and the data from the pets questionnaire you made for the exercise in Section\ \@ref(inputs).

## Your App {#your-app-outputs}

In the app you're developing, add relevant outputs, such as for plots or tables, and the appropriate render function for each output. You can leave the contents blank for now or add in some code to create output.