# Reactive functions {#reactives}

There are a lot of great tutorials that explain the principles behind reactive functions, but that never made any sense to me when I first started out, so I'm just going to give you examples that you can extrapolate principles from.

Reactivity is how Shiny determines which code in `r func("server")` gets to run when. Some types of `r glossary("object", "objects")`, such as the `input` object or objects made by `r func("reactiveValues")`, can trigger some types of functions to run whenever they change. 

For our example, we will use the `reactive_demo` app. It shows three select inputs that allow the user to choose values from the cut, color, and clarity columns of the `diamonds` dataset from `r pkg("ggplot2")`, then draws a plot of the relationship between carat and price for the selected subset.

```{r reactive-demo-app, eval=TRUE, echo=FALSE, fig.cap="Reactive Demo App. You can also access this app with `shinyintro::app(\"reactive_demo\")`or view it in a separate tab with the [showcase interface](https://shiny.psy.gla.ac.uk/debruine/reactive_demo/){target=\"_blank\"}."}
if (is_html)
  knitr::include_app("https://shiny.psy.gla.ac.uk/debruine/reactive_demo/",
height = "800px")
```

Here is the relevant code for the UI. There are four inputs: cut, color, clarity, and update. There are two outputs: title and plot.

```{r}
box(
  title = "Diamonds",
  solidHeader = TRUE,
  selectInput("cut", "Cut", levels(diamonds$cut)),
  selectInput("color", "Color", levels(diamonds$color)),
  selectInput("clarity", "Clarity", levels(diamonds$clarity)),
  actionButton("update", "Update Plot")
),
box(
  title = "Plot",
  solidHeader = TRUE,
  textOutput("title"),
  plotOutput("plot")
)
```


Whenever an input changes, it will trigger some types of functions to run.

## observeEvent()

We learned about `r func("observeEvent")` in Section\ \@ref(first-reactive). This function runs the code whenever the value of the first argument changes. If there are reactives inside the function, they won't trigger the code to run when they change.

```{r}
server <- function(input, output, session) {
  observeEvent(input$update, {
    data <- filter(diamonds,
                   cut == input$cut,
                   color == input$color,
                   clarity == input$clarity)
    
    title <- sprintf("Cut: %s, Color: %s, Clarity: %s",
                     input$cut,
                     input$color,
                     input$clarity)
    
    output$plot <- renderPlot({
      ggplot(data, aes(carat, price)) +
        geom_point(color = "#605CA8", alpha = 0.5) +
        geom_smooth(method = lm, color = "#605CA8")
    })
    
    output$title <- renderText(title)
  })
} 
```

In the example above, which inputs will trigger `r func("renderPlot")` to run and produce a new plot?

`r webex::mcq(c("cut", "color", "clarity", answer = "update", "cut, color & clarity", "all of the above"))`

In the example above, which inputs will trigger `r func("renderText")` to run and produce a new title?

`r webex::mcq(c("cut", "color", "clarity", answer = "update", "cut, color & clarity", "all of the above"))`


## Render functions

Functions that render an output, like `r func("renderText")` or `r func("renderPlot")` will run whenever an input in their code changes. You can trigger a render function just by putting a reactive alone on a line, even if you aren't using it in the rest of the code.

```{r}
server <- function(input, output, session) {
  output$plot <- renderPlot({
    data <- filter(diamonds,
                   cut == input$cut,
                   color == input$color,
                   clarity == input$clarity)
    
    ggplot(data, aes(carat, price)) +
      geom_point(color = "#605CA8", alpha = 0.5) +
      geom_smooth(method = lm, color = "#605CA8")
  })
  
  output$title <- renderText{
    input$update # just here to trigger the function
    
    sprintf("Cut: %s, Color: %s, Clarity: %s",
                     input$cut,
                     input$color,
                     input$clarity)
  })
} 
```

In the example above, which inputs will trigger `r func("renderPlot")` to run and produce a new plot?  
`r webex::mcq(c("cut", "color", "clarity", "update", answer = "cut, color & clarity", "all of the above"))`

Which inputs will trigger `r func("renderText")` to run and produce a new title?  
`r webex::mcq(c("cut", "color", "clarity", "update", "cut, color & clarity", answer = "all of the above"))`

## reactive()

If you move the `data` filtering outside of `r func("renderPlot")`, you'll get an error message like "Can't access reactive value 'cut' outside of reactive consumer." This means that the `input` values can only be read inside certain functions, like `r func("reactive")`, `r func("observeEvent")`, or a render function.

However, we can put the data filtering inside `r func("reactive")`. This means that whenever an input inside that function changes, the code will run and update the value of `r func("data")`. This can be useful if you need to recalculate the data table each time the inputs change, and then use it in more than one function.

```{r}
server <- function(input, output, session) {
  data <- reactive({
    filter(diamonds,
           cut == input$cut,
           color == input$color,
           clarity == input$clarity)
  })
  
  title <- reactive({
    sprintf("Cut: %s, Color: %s, Clarity: %s, N: %d",
                     input$cut,
                     input$color,
                     input$clarity,
            nrow(data()))
  })
  
  output$plot <- renderPlot({
    ggplot(data(), aes(carat, price)) +
      geom_point(color = "#605CA8", alpha = 0.5) +
      geom_smooth(method = lm, color = "#605CA8")
  })
  
  output$text <- renderText(title())
} 
```

In the example above, which inputs will trigger `r func("renderPlot")` to run and produce a new plot?  
`r webex::mcq(c("cut", "color", "clarity", "update", answer = "cut, color & clarity", "all of the above"))`

Which inputs will trigger `r func("renderText")` to run and produce a new title?  
`r webex::mcq(c("cut", "color", "clarity", "update", answer = "cut, color & clarity", "all of the above"))`

::: {.warning}
My most common error is trying to use `data` or `title` as an object instead of as a function. Notice how the first argument to ggplot is no longer `data`, but `r func("data")` and you set the value of data with `data(newdata)`, not `data <- newdata`. For now, just remember this as a quirk of shiny.
:::

## eventReactive() 

While `r func("reactive")` is triggered whenever any input values inside it change, `r func("eventReactive")` is only triggered when the value of the first argument changes, like `r func("observeEvent")`, but returns a reactive function like `r func("reactive")`.

```{r}
server <- function(input, output, session) {
  data <- eventReactive(input$update, {
    filter(diamonds,
           cut == input$cut,
           color == input$color,
           clarity == input$clarity)
  })
  
  title <- eventReactive(input$update, {
    sprintf("Cut: %s, Color: %s, Clarity: %s",
                     input$cut,
                     input$color,
                     input$clarity)
  })
  
  output$plot <- renderPlot({
    ggplot(data(), aes(carat, price)) +
      geom_point(color = "#605CA8", alpha = 0.5) +
      geom_smooth(method = lm, color = "#605CA8")
  })
  
  output$text <- renderText(title())
}
```

In the example above, which inputs will trigger `r func("renderPlot")` to run and produce a new plot?

`r webex::mcq(c("cut", "color", "clarity", answer = "update", "cut, color & clarity", "all of the above"))`

Which inputs will trigger `r func("renderText")` to run and produce a new title?

`r webex::mcq(c("cut", "color", "clarity", answer = "update", "cut, color & clarity", "all of the above"))`


## reactiveVal()

Another pattern that accomplishes a similar thing to `r func("reactive")` is to create a reactive value using `r func("reactiveVal")`. This allows you to update the value of `r func("data")` not just using the code inside the `r func("reactive")` function that created it, but in any function. This is useful when you have multiple functions that need to update that value.

Here, we use `r func("observeEvent")` to trigger the data filtering code only when the update button is pressed. This new data set is assigned to `r func("data")` using the code `data(newdata)`. 

Because `r func("data")` returns a reactive value, it will trigger `r func("renderPlot")` whenever it changes.

```{r}
server <- function(input, output, session) {
    data <- reactiveVal(diamonds)
    
    observeEvent(input$update, {
        newdata <- filter(diamonds,
               cut == input$cut,
               color == input$color,
               clarity == input$clarity)
        
        data(newdata) # updates data()
    })
    
    output$plot <- renderPlot({
        ggplot(data(), aes(carat, price)) +
            geom_point(color = "#605CA8", alpha = 0.5) +
            geom_smooth(method = lm, color = "#605CA8")
    })
    
    output$title <- renderText({
      sprintf("Cut: %s, Color: %s, Clarity: %s",
                     input$cut,
                     input$color,
                     input$clarity)
    })
} 
```

In the example above, which inputs will trigger `r func("renderPlot")` to run and produce a new plot?  
`r webex::mcq(c("cut", "color", "clarity", answer = "update", "cut, color & clarity", "all of the above"))`

Which inputs will trigger `r func("renderText")` to run and produce a new title?  
`r webex::mcq(c("cut", "color", "clarity", "update", answer = "cut, color & clarity", "all of the above"))`

::: {.info}
We used `data <- reactiveVal(diamonds)` in order for `r func("data")` to have a value that didn't cause an error when `r func("renderPlot")` runs for the first time. 
:::

## reactiveValue()

You need to set up a new `r func("reactiveVal")` for each value in an app that you want to make reactive. I prefer to use `r func("reactiveValues")` because it can be used for any new reactive value you need and works just like `input`. 

You can just set your new object to `r func("reactiveValues")` or you can initialise it with starting values like below. The object `v` is a named list, just like `input`, and when it's values change, it triggers reactive functions exactly like `input` does. 

```{r}
server <- function(input, output, session) {
    v <- reactiveValues(
        data = diamonds,
        title = "All Data"
    )
    
    observeEvent(input$update, {
        v$data <- filter(diamonds,
                       cut == input$cut,
                       color == input$color,
                       clarity == input$clarity)
        
        v$title <- sprintf("Cut: %s, Color: %s, Clarity: %s",
                         input$cut,
                         input$color,
                         input$clarity)
    })
    
    output$plot <- renderPlot({
        ggplot(v$data, aes(carat, price)) +
            geom_point(color = "#605CA8", alpha = 0.5) +
            geom_smooth(method = lm, color = "#605CA8")
    })
    
    output$title <- renderText(v$title)
} 
```

In the example above, which inputs will trigger `r func("renderPlot")` to run and produce a new plot?  
`r webex::mcq(c("cut", "color", "clarity", answer = "update", "cut, color & clarity", "all of the above"))`

Which inputs will trigger `r func("renderText")` to run and produce a new title?  
`r webex::mcq(c("cut", "color", "clarity", answer = "update", "cut, color & clarity", "all of the above"))`

::: {.warning}
Note that you refer to reactive values set up this way as `v$data` and `v$title`, not `r func("data")` and `r func("title")`.
:::

## isolate()

If you want to use an input or reactive value inside a reactive function, but don't want to trigger that function, you can `r func("isolate")` it. You can also use `r func("isolate")` to get a reactive value outside a reactive function.

```{r}
server <- function(input, output, session) {
  data <- reactive({
    filter(
      diamonds,
      cut == isolate(input$cut),
      color == isolate(input$color),
      clarity == input$clarity
    )
  })
  
  title <- reactive({
    sprintf(
      "Cut: %s, Color: %s, Clarity: %s",
      input$cut,
      isolate(input$color),
      isolate(input$clarity)
    )
  })
  
  # what is the title at initialisation?
  debug_msg(isolate(title()))
  
  output$plot <- renderPlot({
    ggplot(data(), aes(carat, price)) +
      geom_point(color = "#605CA8", alpha = 0.5) +
      geom_smooth(method = lm, color = "#605CA8")
  })
  
  output$title <- renderText(title())
} 
```

In the example above, which inputs will trigger `r func("renderPlot")` to run and produce a new plot?  
`r webex::mcq(c("cut", "color", answer = "clarity", "update", "cut, color & clarity", "all of the above"))`

Which inputs will trigger `r func("renderText")` to run and produce a new title?  
`r webex::mcq(c(answer = "cut", "color", "clarity", "update", "cut, color & clarity", "all of the above"))`


## Further Resources {#resources-reactive)

* [Mastering Shiny - Basic Reactivity](https://mastering-shiny.org/basic-reactivity.html){target="_blank"}
* [Reactivity - An overview](https://shiny.rstudio.com/articles/reactivity-overview.html){target="_blank"}
* [ Use reactive expressions ](https://shiny.rstudio.com/tutorial/written-tutorial/lesson6/){target="_blank"}
* [Mastering Shiny - Mastering Reactivity](https://mastering-shiny.org/reactivity-intro.html){target="_blank"}



## Exercises {#exercises-reactive}

For the following exercises, clone "reactive_demo" and replace the boxes in the ui with the code below. Delete all the code in `r func("server")`. Make sure this runs before you go ahead.

```{r}
box(width = 4,
    selectInput("stat", "Statistic", c("mean", "sd")),
    selectInput("group", "Group By", c("vore", "order", "conservation")),
    actionButton("update", "Update Table")),
box(width = 8,
    solidHeader = TRUE,
    title = textOutput("caption"),
    tableOutput("table"))
```

You will grouping and summarising the `msleep` data table from `r pkg("ggplot2")` by calculating the mean or standard deviation for all (or some) of the numeric columns grouped by the categorical columns `vore`, `order`, or `conservation`. If you're not sure how to create such a summary table with `r pkg("dplyr")`, look at the following code for a concrete example.

`r hide("Hint")`
```{r, eval = TRUE}
msleep %>%
  group_by(vore) %>%
  summarise_if(is.numeric, "mean", na.rm = TRUE)
```
`r unhide()`

### observeEvent

Use `r func("observeEvent")` to update the output table with the appropriate summary table and to update the caption with an appropriate caption only when the update button is clicked.

`r hide()`
```{r}
server <- function(input, output, session) {
  observeEvent(input$update, {
    data <- msleep %>%
      group_by(.data[[input$group]]) %>%
      summarise_if(is.numeric, input$stat, na.rm = TRUE)
    output$table <- renderTable(data)
    
    caption <-
      sprintf("%ss by %s", toupper(input$stat), input$group)
    output$caption <- renderText(caption)
  })
} 
```
`r unhide()`

### render

Use render functions to update the output table and caption whenever group or stat change.

`r hide()`
```{r}
server <- function(input, output, session) {
  output$table <- renderTable({
    msleep %>%
      group_by(.data[[input$group]]) %>%
      summarise_if(is.numeric, input$stat, na.rm = TRUE)
  })
  
  output$caption <- renderText({
    sprintf("%ss by %s", toupper(input$stat), input$group)
  })
} 
```
`r unhide()`

### reactive

Use `r func("reactive")` to update the output table and caption whenever group or stat change. Ignore the update button.

`r hide()`
```{r}
server <- function(input, output, session) {
  data <- reactive({
    msleep %>%
      group_by(.data[[input$group]]) %>%
      summarise_if(is.numeric, input$stat, na.rm = TRUE)
  })
  output$table <- renderTable(data())
  
  caption <- reactive({
    sprintf("%ss by %s", toupper(input$stat), input$group)
  })
  output$caption <- renderText(caption())
} 
```
`r unhide()`

### eventReactive

Use `r func("eventReactive")` to update the output table and caption only when the update button is clicked.

`r hide()`
```{r}
server <- function(input, output, session) {
  data <- eventReactive(input$update, {
    msleep %>%
      group_by(.data[[input$group]]) %>%
      summarise_if(is.numeric, input$stat, na.rm = TRUE)
  })
  output$table <- renderTable(data())
  
  caption <- eventReactive(input$update, {
    sprintf("%ss by %s", toupper(input$stat), input$group)
  })
  output$caption <- renderText(caption())
} 
```
`r unhide()`


### reactiveVal

Use `r func("reactiveVal")` to update the output table and caption only when the update button is clicked.

`r hide()`
```{r}
server <- function(input, output, session) {
  data <- reactiveVal()
  caption <- reactiveVal()
  
  observeEvent(input$update, {
    newdata <- msleep %>%
      group_by(.data[[input$group]]) %>%
      summarise_if(is.numeric, input$stat, na.rm = TRUE)
    data(newdata)
    
    # this is an alternative way to set reactiveVal
    # by piping the value into the function
    sprintf("%ss by %s", toupper(input$stat), input$group) %>%
      caption()
  })
  
  output$table <- renderTable(data())
  output$caption <- renderText(caption())
} 
```
`r unhide()`


### reactiveValues

Use `r func("reactiveValues")` to update the output table and caption only when the update button is clicked.

`r hide()`
```{r}
server <- function(input, output, session) {
  v <- reactiveValues()
  
  observeEvent(input$update, {
    v$data <- msleep %>%
      group_by(.data[[input$group]]) %>%
      summarise_if(is.numeric, input$stat, na.rm = TRUE)
    
    v$caption <-
      sprintf("%ss by %s", toupper(input$stat), input$group)
  })
  
  output$table <- renderTable(v$data)
  output$caption <- renderText(v$caption)
} 
```
`r unhide()`

## Your App {#your-app-reactive}

Add reactive functions to your custom app. Think about which patterns are best for your app. For example, if you need to update a data table when inputs change, and then use it in more than one output, it's best to use `r func("reactive")` to create a function for the data and callit in the render functions for each output, rather than creating the data table in each render function.

