# Contingent Display {#contingency}

```{r contingency-demo-app, eval=TRUE, echo=FALSE, fig.cap="Contingency Demo App. You can also access this app with `shinyintro::app(\"contingency_demo\")`or view it in a separate tab with the [showcase interface](https://shiny.psy.gla.ac.uk/debruine/contingency_demo/){target=\"_blank\"}."}
if (is_html)
  knitr::include_app("https://shiny.psy.gla.ac.uk/debruine/contingency_demo/",
height = "600px")
```

## Hide and Show

I frequently want to make some aspect of a shiny app contingent on the state of another aspect, such as only showing a text input of the value of a select input is "other". You can use the `r func("hide")` and `r func("show")` functions from `r pkg("shinyjs")` to accomplish this easily.

When you set up the UI, wrap any elements that should be hidden at the start in `r func("hidden")`.

```{r}
# in the ui
box(title = "Questions",
    solidHeader = TRUE,
    selectInput("first_pet", "What was your first pet?", 
                c("", "dog", "cat", "ferret", "other")),
    hidden(textInput("first_pet_other", NULL, 
                     placeholder = "Specify the other pet"))
)
```

Then set up the hide and show logic in `r func("server")`. 

```{r}
# in the server
observeEvent(input$first_pet, {
    if (input$first_pet == "other") {
        show("first_pet_other")
    } else {
        hide("first_pet_other")
    }
})
```

### Groups

Sometimes you need to hide and show a group of elements, depending on something else. You can wrap the grouped elements in a div tag with an id and hide and show that id.

For example, it doesn't make sense to show the questions above to someone who has never had a pet. Add a `r func("selectInput")` before the previous two questions, and then wrap those questions in `r func("tags$div")` with an id of `r dt("first_pet_grp")`

```{r}
# replace in ui
box(
  title = "Questions",
  solidHeader = TRUE,
  selectInput("had_pet", "Have you ever had a pet?", c("", "Yes", "No")),
  hidden(tags$div(
    id = "first_pet_grp",
    selectInput("first_pet", "What was your first pet?",
                c("", "dog", "cat", "ferret", "other")),
    textInput("first_pet_other", NULL,
              placeholder = "Specify the other pet")
  ))
)
```

Then add the following code to the server function to hide or show `first_pet_grp` depending on the value of `had_pet`. The server code above will take care of whether or not `first_pet_other` is visible.

```{r}
# add to server
observeEvent(input$had_pet, {
  if (input$had_pet == "Yes") {
    show("first_pet_grp")
  } else {
    hide("first_pet_grp")
  }
})
```

::: {.try data-latex=""}
Try to figure out what could go wrong if you didn't wrap "first_pet" and "first_pet_other" in a group, and instead just hid or showed "first_pet" and "first_pet_other" depending on the value of `has_pet`?
:::

### Toggle

Sometimes you need to change the visibility of an element when something happens, rather than specifically hide or show it. You can use `r func("toggle")` to hide an element if it's visible and show it if it's hidden.

Add an `r func("actionButton")` to the sidebar menu (not inside the box) and give the box an `r arg("id")` of `r dt("pet_box")`. Any element that you might want to refer to in the code needs an `r arg("id")`.

```{r}
# add to ui
actionButton("toggle_pet_box", "Toggle Pet Questions")
```

Now, whenever you click the "toggle_pet_box" button, the contents of "pet_box" will toggle their visibility.

```{r}
# add to server
observeEvent(input$toggle_pet_box, {
  toggle("pet_box")
})
```

::: {.try data-latex=""}
What would go wrong if you put the button inside the box?
:::

## Changing Styles

You can use `r func("addClass")`, `r func("removeClass")`, and `r func("toggleClass")` to change element classes. You usually want to do this with classes you've defined yourself. 

Add the following style to the `r path("www/custom.css")` file.

```{r}
.notice-me {
  color: red;
  text-decoration: underline;
  font-weight: 800;
}
```

And add this box to the ui:

```{r}
box(title = "Notice", solidHeader = TRUE,
    p(id = "notice_text", "Please pay attention to this text."),
    actionButton("add_notice", "Notice Me"),
    actionButton("remove_notice", "Ignore Me"),
    actionButton("toggle_notice", "Toggle Me")
)
```

This code adds the class `r dt("notice-me")` to the paragraph element `r dt("notice_text")` whenever the "add_notice" button is pressed.

```{r}
observeEvent(input$add_notice, {
  addClass("notice_text", "notice-me")
})
```

::: {.try data-latex=""}
Guess how you would use `r func("removeClass")`, and `r func("toggleClass")` with the buttons set up above.
:::


### Changing non-shiny elements

Unfortunately, not all elements on the web page have an ID that can be altered by `r func("addClass")` or `r func("removeClass")`. For example, the skin of a shinydashboard app is determined by the css class of the body element. However, we can use `r func("runjs")` to run any arbitrary [JavaScript](#JavaScript) code. 

Add the following action button into the `r func("sidebarMenu")`.

```{r}
actionButton("random_skin", "Random Skin")
```

The `r glossary("jQuery")` code below changes the skin of your app on a button press by removing all possible skin-color classes and adding a random one.

```{r}
observeEvent(input$random_skin, {
  skins <- c("red", "yellow", "green", "blue", "purple", "black")
  skin_color <- sample(skins, 1)
  
  js <- sprintf("$('body').removeClass('%s').addClass('skin-%s');",
                paste(paste0("skin-", skins), collapse = " "),
                skin_color)
  
  shinyjs::runjs(js)
})
```

::: {.info data-latex=""}
Changing the skin color with a button press isn't something you'll easily find documented in online materials. I figured it out through looking at how the underlying html changed when I changed the skin color in the app code. Hacks like this require lots of trial and error, but get easier the more you understand about html, css and JavaScript.
:::

## Changing input options

The relevant options in a `r func("selectInput")` or `r func("radioButton")` may change depending on the values of other inputs. Sometimes you can accommodate this by creating multiple versions of a input and hiding or showing. Other times you may wish to update the input directly.

Add the following box to the ui.

```{r}
box(title = "Data", solidHeader = TRUE, width = 12,
  selectInput("dataset", "Choose a dataset", c("mtcars", "sleep")),
  checkboxGroupInput("columns", "Select the columns to show", inline = TRUE),
  tableOutput("data_table")
)
```

First, set up the code to display the correct data in the table.

```{r}
mydata <- reactive({
  get(input$dataset, "package:datasets")
})

output$data_table <- renderTable(mydata())
```


Now we need to set the options for `r dt("columns")` depending on which "dataset" is selected.

```{r}
observe({
  col_names <- names(data())
  debug_msg(col_names)
  updateCheckboxGroupInput(inputId = "columns",
                           choices = col_names,
                           selected = col_names)
})
```

Finally, we can add some code to select only the checked columns to display.

```{r}
observe({
  full_data <- get(input$dataset, "package:datasets")
  col_names <- names(full_data)
  updateCheckboxGroupInput(
    inputId = "columns",
    choices = col_names,
    selected = col_names,
    inline = TRUE
  )
})
```

::: {.try data-latex=""}
Why do we have to get the dataset again instead of using the data from `r func("mydata")`?
:::

Finally, alter the reactive function to only show the selected columns.

```{r}
mydata <- reactive({
  d <- get(input$dataset, "package:datasets")
  d[input$columns]
})
```

::: {.try data-latex=""}
What happens when you unselect all the columns? How can you fix this?
:::

## Further Resources {#resources-contingency}

* [Mastering Shiny: Packages: Dynamic UI](https://mastering-shiny.org/action-dynamic.html){target="_blank"}

## Exercises {#exercises-contingency}

### Filtered data  {-}

Create an app where you use inputs to filter a dataset and display a table of the filtered dataset. For example, with the `msleep` dataset, you could have inputs that select `vore`, `order` and `conservation`. Since some values will exclude categories (e.g., there are no omnivores in the order Cetacea), update the available categories in each input when values are selected. Make sure you have a way to reset the values.

## Your App {#your-app-contingency}

Check for places in your app that could use contingency.

