# Customized reports {#reports}

While the best part of Shiny apps is their interactivity, sometimes your users need to download data, images, or a static report. This section will show you how.

```{r reports-demo-app, eval=TRUE, echo=FALSE, fig.cap="Reports Demo App. You can also access this app with `shinyintro::app(\"reports_demo\")`or view it in a separate tab with the [showcase interface](https://shiny.psy.gla.ac.uk/debruine/reports_demo/){target=\"_blank\"}."}
if (is_html)
  knitr::include_app("https://shiny.psy.gla.ac.uk/debruine/reports_demo/",
                     height = "475px"
  )
```


## Download Data

First, we need to add the appropriate UI to our questionnaire app. Create a new tab called `r dt("report_tab")` with two `r func("downloadButton")`s, one for the pets data and one for the food data.

```{r}
report_tab <- tabItem(
  tabName = "report_tab",
  box(
    id = "download_box",
    title = "Downloads",
    solidHeader = TRUE,
    downloadButton("pet_data_dl", "Pets Data"),
    downloadButton("food_data_dl", "Food Data")
  )
)
```

::: {.warning data-latex=""}
Remember to add this to `r func("tabItems")` in `r func("dashboardBody")` and also add a corresponding `r func("menuItem")` in `r func("sidebarMenu")`. Run the app to make sure the UI looks like you expect before you proceed.
:::

Now we need to add code to `r func("server")` to handle the downloads. `r func("downloadButton")` is a special type of output that is handled by `r func("downloadHandler")`. This function takes two arguments, a function to create the `r arg("filename")` and a function to create the `r arg("content")`. 

```{r}
### pet_data_dl ----
output$pet_data_dl <- downloadHandler(
  filename = function() {
    debug_msg("pet_data_dl")
    paste0("pet-data_", Sys.Date(), ".csv")
  },
  content = function(file) {
    gs4_write_csv(v$pet_summary_data, file)
  }
)
```

::: {.warning data-latex=""}
Instead of `r func("write.csv")` or `r func("readr::write_csv")`, here we're using `r func("gs4_write_csv")` (defined in `r path("scripts/gs4.R")`) because googlesheets can return list columns that cause errors when [saving to CSV](#gs4_save) without some preprocessing.
:::


## Download Images

Now that we need to use the summary plot in more than one place, it doesn't make sense to build it twice. Move all of the code from the `r func("renderPlot")` for `output$pet_summary` into a `r func("reactive")` called `pet_summary_plot`. Then we can build the plot whenever the inputs change and refer to it anywhere as `r func("pet_summary_plot")`.

```{r}
### pet_summary_plot ----
pet_summary_plot <- reactive({ debug_msg("pet_summary_plot")
  # code from output$pet_summary ...
})

### pet_summary ----
output$pet_summary <- renderPlot({ debug_msg("pet_summary")
    pet_summary_plot()
})
```

The `r func("downloadHandler")` works the same as for downloading a CSV file. You can use `r func("ggsave")` to write the plot. 

```{r}
# pet_plot_dl ----
output$pet_plot_dl <- downloadHandler(
  filename = function() {
    paste0("pet-plot_", Sys.Date(), ".png")
  },
  content = function(file) {
    ggsave(file,
           pet_summary_plot(),
           width = 7,
           height = 5)
  }
)
```

::: {.try data-latex=""}
Add numeric inputs to the UI to let the user specify the downloaded plot `r arg("width")` and `r arg("height")`.

`r hide()`
```{r}
# add to report_tab ui
numericInput("plot_width", "Plot Width (inches)", 7, min = 1, max = 10)
numericInput("plot_height", "Plot Height (inches)", 5, min = 1, max = 10)

### pet_plot_dl ----
output$pet_plot_dl <- downloadHandler(
  filename = function() {
    debug_msg("pet_plot_dl")
    paste0("pet-plot_", Sys.Date(), ".png")
  },
  content = function(file) {
    ggsave(
      file,
      pet_summary_plot(),
      width = input$plot_width,
      height = input$plot_height
    )
  }
)

```
`r unhide()`
:::


## R Markdown

You can render an `r glossary("R Markdown")` report for users to download.

First, you need to save an R Markdown file (save the one below as `r path("reports/report.Rmd")`). You need to set up `params` for any info that you want to pass from the Shiny app in th `r glossary("YAML")` header. Here, we will dynamically update the title, data, and plot. You can set the values to `NULL` or another default value. 

``` yml
---
title: "`r knitr::inline_expr("params$title")`"
date: "`r knitr::inline_expr("format(Sys.time(), '%d %B, %Y')")`" 
output: html_document
params:
  title: Report
  data: NULL
  plot: NULL
---
```

Then you can refer to these params in the R Markdown file as `params$name`. If you knit the report, these will take on the default values.

````
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)

library(dplyr)
```

```{r}
n_responses <- nrow(params$data)

n_sessions <- params$data$session_id %>%
  unique() %>%
  length()
```

We asked people to rate how much they like pets. We have obtained `r knitr::inline_expr("n_responses")` responses from `r knitr::inline_expr("n_sessions")` unique sessions.

## Summary Plot

```{r plot}
params$plot
```
````

The `content` function uses `r func("rmarkdown::render")` to create the output file from Rmd. We are using `html_document` as the output type, but you can also render as a PDF or Word document if you have `r glossary("pandoc")` or `r glossary("latex")` installed. Include the `r arg("title")`, `r arg("data")` and `r arg("plot")` as a named list to the `r arg("params")` argument. 

```{r}
### pet_report_dl ----
output$pet_report_dl <- downloadHandler(
  filename = function() {
    debug_msg("pet_report_dl")
    paste0("pet-report_", Sys.Date(), ".html")
  },
  content = function(file) {
    rmarkdown::render("reports/report.Rmd",
                      output_file = file, 
                      params = list(
                        title = "Pet Report", 
                        data = v$pet_summary_data,
                        plot = pet_summary_plot()
                      ),
                      envir = new.env(),
                      intermediates_dir = tempdir())
  }
)
```

Setting the `r arg("envir")` argument to `r func("new.env")` makes sure that settings in your Shiny app, such as a default ggplot theme, can't affect the rendered file, but also makes it so that the R Markdown code will not have access to any objects from your app, such as `input` or `v$summary_data`, unless you pass them as parameters.

::: {.dangerous data-latex=""}
On your own computer, you don't need to set `intermediates_dir = tempdir()`, but you will need to do this if you want to deploy your app on a shiny server. When rmarkdown renders an Rmd file, it creates several intermediate files in the working directory and then deletes them. you have permission to write to the app's working directory on your own computer, but might not on a shiny server. `r func("tempdir")` is almost always a safe place to write temporary files to.
:::

## Further Resources {#resources-reports}

* [Mastering Shiny: Uploads and Downloads](https://mastering-shiny.org/action-transfer.html){target="_blank"}
* [R Markdown](https://rmarkdown.rstudio.com/){target="_blank"}
* [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/){target="_blank"}



## Exercises {#exercises-reports}

### Food Data {-}

Create a button that downloads the food data.

`r hide()`
```{r}
# add to report_tab ui ----
downloadButton("food_data_dl", "Food Data")

### food_data_dl ----
output$food_data_dl <- downloadHandler(
  filename = function() {
    debug_msg("food_data_dl")
    paste0("food-data_", Sys.Date(), ".csv")
  },
  content = function(file) {
    gs4_write_csv(v$food_summary_data, file)
  }
)
```
`r unhide()`

### Food Image {-}

Create a button that downloads the food summary plot

`r hide()`
```{r}
# add to report_tab ui ----
downloadButton("food_plot_dl", "Food Plot")

### food_summary_plot ----
food_summary_plot <- reactive({ debug_msg("food_summary_plot")
  # code from output$food_summary ...
})

### food_summary ----
output$food_summary <- renderPlot({ debug_msg("food_summary")
  food_summary_plot()
})

### food_plot_dl ----
output$food_plot_dl <- downloadHandler(
  filename = function() {
    debug_msg("food_plot_dl")
    paste0("food-plot_", Sys.Date(), ".png")
  },
  content = function(file) {
    ggsave(
      file,
      food_summary_plot(),
      width = input$plot_width,
      height = input$plot_height
    )
  }
)
```
`r unhide()`

### Food Report {-}

Create a downloadable report for the food data.

`r hide()`
```{r}
# add to report_tab ui ----
downloadButton("food_report_dl", "Food Report")

### food_report_dl ----
output$food_report_dl <- downloadHandler(
  filename = function() {
    debug_msg("food_report_dl")
    paste0("food-report_", Sys.Date(), ".html")
  },
  content = function(file) {
    rmarkdown::render("reports/report.Rmd",
                      output_file = file, 
                      params = list(
                        title = "Food Report", 
                        data = v$food_summary_data,
                        plot = food_summary_plot()
                      ),
                      envir = new.env(),
                      intermediates_dir = tempdir())
  }
)
```
`r unhide()`

## Your app {#your-app-reports}

What might users want to download from your app? Include data, image, or report download if that is applicable to your app.
