# Installing `R` {#installing-r}

Installing R and RStudio is usually straightforward. The sections below explain how and [there is a helpful YouTube video here](https://www.youtube.com/watch?v=lVKMsaWju8w).

## Installing Base R

Install base R from <https://cran.rstudio.com/>. Choose the download link for your operating system (Linux, Mac OS X, or Windows).

If you have a Mac, install the latest release from the newest `R-x.x.x.pkg` link (or a legacy version if you have an older operating system). After you install R, you should also install [XQuartz](http://xquartz.macosforge.org/) to be able to use some visualisation packages.

If you are installing the Windows version, choose the "[base](https://cran.rstudio.com/bin/windows/base/)" subdirectory and click on the download link at the top of the page. After you install R, you should also install [RTools](https://cran.rstudio.com/bin/windows/Rtools/); use the "recommended" version highlighted near the top of the list.

If you are using Linux, choose your specific operating system and follow the installation instructions.

## Installing RStudio

Go to [rstudio.com](https://www.rstudio.com/products/rstudio/download/#download) and download the RStudio Desktop (Open Source License) version for your operating system under the list titled **Installers for Supported Platforms**.

### RStudio Settings

There are a few settings you should fix immediately after updating RStudio. Go to **`Global Options...`** under the **`Tools`** menu (&#8984;,), and in the General tab, uncheck the box that says **`Restore .RData into workspace at startup`**.  If you keep things around in your workspace, things will get messy, and unexpected things will happen. You should always start with a clear workspace. This also means that you never want to save your workspace when you exit, so set this to **`Never`**. The only thing you want to save are your scripts.

You may also want to change the appearance of your code. Different fonts and themes can sometimes help with visual difficulties or [dyslexia]((https://datacarpentry.org/blog/2017/09/coding-and-dyslexia){target="_blank"}. I prefer a dark background and the FiraCode font for my own coding, but I will use the default theme for screenshots in this book.

```{r settings-general, eval = TRUE, echo=FALSE, out.width = "45%", fig.cap = "My RStudio General and Appearance settings"}
knitr::include_graphics(
  c('images/rstudio_settings_general.png', 
    'images/rstudio_settings_appearance.png'))
```

You may also want to change the settings in the Code tab. I prefer two spaces instead of tabs for my code and I like to be able to see the `r glossary("whitespace")` characters. But these are all a matter of personal preference.

```{r settings-code, eval = TRUE, echo=FALSE, out.width = "45%", fig.cap = 'My RStudio Code settings'}
knitr::include_graphics(
  c("images/rstudio_settings_code_editing.png",
    "images/rstudio_settings_code_display.png"))
```

### Installing shinyintro

To install the class package, which will provide you with a copy of all of the shiny apps we'll use for demos and the basic template, paste the following code into the console in RStudio.

```{r}
# you may have to install devtools first with 
# install.packages("devtools")

devtools::install_github("debruine/shinyintro")
```


## Installing LaTeX

You can install the LaTeX typesetting system to produce PDF reports from RStudio. Without this additional installation, you will be able to produce reports in HTML but not PDF. This course will not require you to make PDFs. To generate PDF reports, you will additionally need: 

1.  [pandoc](http://pandoc.org/installing.html), and
2.  LaTeX, a typesetting language, available for
    -   WINDOWS: [MikTeX](http://miktex.org/)
    -   Mac OS: [MacTex](https://tug.org/mactex/downloading.html) (3.2GB download) or [BasicTeX](http://ww.tug.org/mactex/morepackages.html) (78MB download, but should work fine)
    -   Linux: [TeX Live](https://www.tug.org/texlive/)

